---
title: "met_lung_sumExp"
author: "Francesc Fernandez"
date: "2/12/2019"
output: html_document
---

#Summary

This script converts the lung Metabolomics dataset we received from Metabolon and saves the relevant information as a SummarizedExperiment object for later use. The main steps include:     
     
* Importing and parsing the data and metadata from the original xls file
* PCA plot of the raw data using the top 100 most variable metabolites
* NA imputation
* PCA plot of the data (log + NA imputed)
* Linear models with interaction
* PLS model
* Mapping metabolites to KEGG pathways
* Export results to SummarizedExperiment object


#Data processing
##Loading libraries

```{r setup, include=FALSE}
library(xlsx)
library(pcaMethods)
library(ggplot2)
library(pls)
library(pheatmap)
library(KEGGREST)
library(GGally)
library(magrittr)

config <- new.env()
source("config.R", config)

knitr::opts_chunk$set(echo = TRUE)
```

##Loading data

```{r, loading metabolomics data}

data.met.lung.raw <- read.xlsx(config$file.metab.lung, sheetName = "OrigScale")

```

##Parsing metadata

```{r, extracting the data read-outs and the metadata for the lung}
meta.data.met.lung <- data.frame(t(data.met.lung.raw[1:14, -c(1:12)]))
colnames(meta.data.met.lung) <- as.matrix(meta.data.met.lung[1,])
meta.data.met.lung <- meta.data.met.lung[-1, ]
meta.data.met.lung$AGE <- factor(meta.data.met.lung$AGE, levels = c("3M", "21M"))
meta.data.met.lung$TRT <- factor(meta.data.met.lung$TRT, levels = c("CTRL", "Bleo"))



data.met.lung.varInfo <- data.met.lung.raw[-c(1:14), ][, 1:13]
colnames(data.met.lung.varInfo) <- as.character(as.matrix(data.met.lung.varInfo[1, ]))  
data.met.lung.varInfo <- data.met.lung.varInfo[-1, ]
data.met.lung.varInfo <- data.frame(metabolite.id = paste0("MetLung", 1:nrow(data.met.lung.varInfo)),
                                    data.met.lung.varInfo)
rownames(data.met.lung.varInfo) <- data.met.lung.varInfo$metabolite.id




data.met.lung <- data.met.lung.raw[-c(1:15), ][, -c(1:13)]
data.met.lung <- apply(data.met.lung, 2, as.numeric)
rownames(data.met.lung) <- data.met.lung.varInfo$metabolite.id


```

##PCA model

```{r}

ind.var.met.lung.top100 <- head(order(apply(t(data.met.lung), 2, var), decreasing = TRUE), 100)

mod.pca.lung.top100.raw <- pca(t(data.met.lung)[, ind.var.met.lung.top100],
                               scale = "uv",
                               nPcs = 10,
                               cv = "q2")


scores.pca.lung.top100.raw <- data.frame(pcaMethods::scores(mod.pca.lung.top100.raw)[, 1:3],
                                     TRT = meta.data.met.lung$TRT,
                                     AGE = meta.data.met.lung$AGE)

ggplot(scores.pca.lung.top100.raw, aes(x = PC1, y = PC2, colour = TRT, shape = AGE)) + geom_point(size = 2) + scale_shape_manual(values=c(1, 16)) + theme_minimal() + ggtitle("Raw PCA plot")

```

##NA imputation

NAs are imputed in a per metabolite basis. We use the minimum value for each metabolite to impute the NAs. We also take the log2 of the intensities after imputation.

```{r}

dat.met.lung <- t(data.met.lung)

v.nas <- is.na(dat.met.lung) %>%
  colMeans %>%
  sort 
v.keep <- names(v.nas)[v.nas < .5]

dat.met.lung.imp <- apply(dat.met.lung[, v.keep], 2, function(x){
                        replace(x, is.na(x), min(na.omit(x)))
}) %>% t


dat.met.lung.imp.log <- log2(1 +  dat.met.lung.imp)


```



```{r}

ind.var.met.lung.top100.log <- head(order(apply(t(dat.met.lung.imp.log), 2, var), decreasing = TRUE), 100)

mod.pca.lung.top100.log <- pca(t(dat.met.lung.imp.log)[, ind.var.met.lung.top100.log],
                           scale = "uv",
                           nPcs = 10,
                           cv = "q2")
scores.pca.lung.top100.log <- data.frame(pcaMethods::scores(mod.pca.lung.top100.log)[, 1:3],
                                     TRT = meta.data.met.lung$TRT,
                                     AGE = meta.data.met.lung$AGE)

ggplot(scores.pca.lung.top100.log, aes(x = PC1, y = PC2, colour = TRT, shape = AGE)) + geom_point(size = 2) + scale_shape_manual(values=c(1, 16)) + theme_minimal() + ggtitle("Raw PCA plot")

```

## Making PLS models with the imputed data

The objective here was to see whether we can use the metabolomics data to separate the 4 groups (Bleo/NaCl and old/young). We also wanted to find out if there are specific patterns in the data that are specific of one or a subgroup of the 4 classes (e.g. is there a pattern when goign from Bleo to NaCL that is only found in old animals or viceversa?)


```{r}

model.matrix.lung.met <- model.matrix( ~ TRT * AGE, 
                                       data = meta.data.met.lung)[, -1]

mod.pls.lung.raw <- plsr(model.matrix.lung.met ~ t(dat.met.lung.imp.log),
                         scale = TRUE,
                         validation = "LOO",
                         jackknife = TRUE,
                         ncomp = 10)

mod.pls.lung.int <- plsr(model.matrix( ~ TRT : AGE, data = meta.data.met.lung)[, which(apply(apply(model.matrix( ~ TRT : AGE, data = meta.data.met.lung), 2, diff),2, function(x){length(unique(x))}) > 1)] ~ t(dat.met.lung.imp.log),
                         scale = TRUE,
                         validation = "LOO",
                         jackknife = TRUE,
                         ncomp = 10)

mod.pls.lung.scores <- as.data.frame(pls::scores(mod.pls.lung.raw)[, 1:3])
colnames(mod.pls.lung.scores) <- c("comp1", "comp2", "comp3")
mod.pls.lung.scores.int <- as.data.frame(pls::scores(mod.pls.lung.int)[, 1:3])
colnames(mod.pls.lung.scores.int) <- c("comp1", "comp2", "comp3")

ggpairs(data = mod.pls.lung.scores.int,
        aes(colour = interaction(meta.data.met.lung$TRT,
                                 meta.data.met.lung$AGE))) + theme_minimal()

ggplot(mod.pls.lung.scores.int, aes(x = comp1,
                                    y = comp2,
                                    colour = interaction(meta.data.met.lung$TRT,
                                                         meta.data.met.lung$AGE))) + geom_point()

```


```{r}

# k <- keggGet("hsa04216")
# 
# met.ids.path <- as.character(data.met.lung.varInfo[na.omit(match(names(k[[1]]$COMPOUND), data.met.lung.varInfo$KEGG)), ]$metabolite.id)
# 
# a <- pca(t(dat.met.lung.imp.log[met.ids.path, ]),
#          cv = "q2",
#          scale = "uv",
#          nPcs = 5)

```

## Fitting linear models with interaction term and extracting p-vals and FCs

```{r}

mod.lms <- apply(dat.met.lung.imp.log, MARGIN = 1, function(metabolite){
                  lm(metabolite ~ meta.data.met.lung$AGE * meta.data.met.lung$TRT)
})

mod.lms.anova <- lapply(mod.lms, anova)
mod.lms.pvals <- t(sapply(mod.lms.anova, function(mod){matrix(mod[c("meta.data.met.lung$AGE", "meta.data.met.lung$TRT", "meta.data.met.lung$AGE:meta.data.met.lung$TRT"), ]$Pr, ncol = 1)}))
colnames(mod.lms.pvals) <- c("AGE", "TRT", "AGE.TRT")
mod.lms.pvals.adj <- data.frame(apply(mod.lms.pvals, 2, function(x){p.adjust(x, method = "fdr")}))

mod.lm.coefs <- t(sapply(mod.lms, coef))[,-1]
colnames(mod.lm.coefs) <- paste("coef", sapply(strsplit(colnames(mod.lm.coefs), "$", fixed = TRUE), function(x){if(length(x)>2){"AGE.TRT"}else{x[2]}}), sep = ".")
colnames(mod.lm.coefs) <- gsub("Bleo|21M", "", colnames(mod.lm.coefs))

lFC <- apply(dat.met.lung.imp, 1, function(met){
              FCs.AGE <- aggregate(as.numeric(met), list(meta.data.met.lung$AGE), mean)
              FCs.AGE <- FCs.AGE[FCs.AGE$Group.1 == "21M", "x"] / FCs.AGE[FCs.AGE$Group.1 == "3M", "x"]
              FCs.TRT <- aggregate(as.numeric(met), list(meta.data.met.lung$TRT), mean)
              FCs.TRT <- FCs.TRT[FCs.TRT$Group.1 == "Bleo", "x"] / FCs.TRT[FCs.TRT$Group.1 == "CTRL", "x"]
              
              return(data.frame(lFC.AGE = log2(FCs.AGE), lFC.TRT = log2(FCs.TRT)))
  
  })
lFCl <- apply(dat.met.lung.imp.log, 1, function(met){
              FCs.AGE <- aggregate(as.numeric(met), list(meta.data.met.lung$AGE), mean)
              FCs.AGE <- FCs.AGE[FCs.AGE$Group.1 == "21M", "x"] - FCs.AGE[FCs.AGE$Group.1 == "3M", "x"]
              FCs.TRT <- aggregate(as.numeric(met), list(meta.data.met.lung$TRT), mean)
              FCs.TRT <- FCs.TRT[FCs.TRT$Group.1 == "Bleo", "x"] - FCs.TRT[FCs.TRT$Group.1 == "CTRL", "x"]
              
              return(data.frame(lFCl.AGE = FCs.AGE, lFCl.TRT = FCs.TRT))
  
  })

lFC <- do.call(rbind, lFC)
lFCl <- do.call(rbind, lFCl)

# write.table(file = "lung_treatmentMet_0_05.txt", matrix(as.character(na.omit(data.met.lung.varInfo[rownames(mod.lms.pvals.adj)[which(mod.lms.pvals.adj$TRT <= 0.05)], ]$KEGG)), ncol = 1), row.names = FALSE, col.names = FALSE, quote = FALSE)
# 
# 
# write.table(file = "lung_AgeMet_0_05.csv", matrix(as.character(na.omit(data.met.lung.varInfo[rownames(mod.lms.pvals.adj)[which(mod.lms.pvals.adj$AGE <= 0.05)], ]$KEGG)), ncol = 1), row.names = FALSE, col.names = FALSE, quote = FALSE)
# 
# write.table(file = "/data/cb_projects/user/fernafra/workspace/projects/Bleomycin_Model_all/data/BleoModel/AgeMet_0_1.csv", matrix(as.character(na.omit(data.met.lung.varInfo[rownames(mod.lms.pvals.adj)[which(mod.lms.pvals.adj$AGE <= 0.1)], ]$KEGG)), ncol = 1), row.names = FALSE, col.names = FALSE, quote = FALSE)
# 
# write.table(file = "/data/cb_projects/user/fernafra/workspace/projects/Bleomycin_Model_all/data/BleoModel/treatmentAgeMet_0_05.csv", matrix(as.character(na.omit(data.met.lung.varInfo[rownames(mod.lms.pvals.adj)[which(mod.lms.pvals.adj$TRT.AGE <= 0.05)], ]$KEGG)), ncol = 1), row.names = FALSE, col.names = FALSE, quote = FALSE)


```


```{r, running limma contrasts}

# library(limma)
# meta.data.met.lung.valNam <- meta.data.met.lung
# meta.data.met.lung.valNam$AGE <- as.factor(make.names(meta.data.met.lung.valNam$AGE))
# meta.data.met.lung.valNam$TRT <- factor(meta.data.met.lung.valNam$TRT)
# 
# meta.data.met.lung.valNam$TRT <- relevel(meta.data.met.lung.valNam$TRT, "CTRL")
# meta.data.met.lung.valNam$AGE <- relevel(meta.data.met.lung.valNam$AGE, "X3M")
# 
# meta.data.met.lung.valNam$group <- interaction(meta.data.met.lung.valNam$TRT,
#                                                 meta.data.met.lung.valNam$AGE)
# 
# limma.fit <- lmFit((dat.met.lung.imp.log),
#                    design = model.matrix(~ 0 + group, data = meta.data.met.lung.valNam))
# 
# # colnames(limma.fit$coefficients) <- c("(Intercept)", sapply(colnames(limma.fit)[-1], function(x){substr(x, start = 4, stop = nchar(x))}))
# 
# contrasts.lung <- makeContrasts(contrasts = c("(groupCTRL.X3M + groupCTRL.X21M)/2  - (groupBleo.X3M + groupBleo.X21M)/2", 
#                                                "(groupCTRL.X3M + groupBleo.X3M)/2  - (groupCTRL.X21M + groupBleo.X21M)/2",
#                                                "(groupCTRL.X3M + groupCTRL.X21M) - (groupBleo.X3M + groupBleo.X21M)"),
#                                  levels = c( "groupCTRL.X3M", "groupBleo.X3M", "groupCTRL.X21M", "groupBleo.X21M"))
# 
# 
# contr<- contrasts.fit(fit = limma.fit,
#                       contrasts = contrasts.lung)
# 
# contr.bayes <- eBayes(contr)
```



## Getting KEGG pathway annotation for each metabolite

When a metabolite does not have a known KEGG ID, the reported pathways will be NA

```{r, generating the mapped pathways for mouse, warning=FALSE}

path.mouse <- KEGGREST::keggList("pathway", organism = "mmu")
path.human <- KEGGREST::keggList("pathway", organism = "hsa")
# 
# path.mouse <- data.table::fread(config$file.kegg.mmu.list, header = FALSE)
# path.mouse <- setNames(path.mouse$V1, path.mouse$V2)
# 
# path.human <- data.table::fread(config$file.kegg.mmu.list, header = FALSE) 
# path.human <- setNames(path.human$V1, path.human$V2)

# cpd.human <- KEGGREST::keggList("cpd", organism = "hsa")
# path.ids.mouse <- substr(sapply(strsplit(names(path.mouse), ":"), function(x){x[2]}), start = 4, stop = 8)
# path.ids.human <- substr(sapply(strsplit(names(path.human), ":"), function(x){x[2]}), start = 4, stop = 8)


data.met.lung.varInfo.exp <- data.frame(data.met.lung.varInfo[rownames(dat.met.lung.imp), ],
                                        pval     = mod.lms.pvals,
                                        pval.adj = mod.lms.pvals.adj)


kegg.path.cpd <- keggLink("pathway", "cpd")

kegg.path.cpd <- data.table::fread(config$file.kegg.mmu.cpd, header = FALSE) 
kegg.path.cpd <- setNames(kegg.path.cpd$V1, kegg.path.cpd$V2)

kegg.path.cpd.df <- sapply(data.met.lung.varInfo.exp$KEGG, function(met){
  
  if(!is.na(met)){
    
  paste0("path:mmu", sapply(sapply(strsplit(kegg.path.cpd[grep(met, names(kegg.path.cpd))], ":"), function(x)x[2]), function(y){substr(y, start = 4, stop = nchar(y))}))  
   
    
  }else{
    ""
  }
})

cross.cpd.path <- sapply(kegg.path.cpd.df, function(pl){
  if(pl != ""){
    paste(sapply(strsplit(path.mouse[which(names(path.mouse) %in% pl)], " - "), function(x)x[1]), collapse = "|")
  }else{
    NA
  }
})

cross.cpd.id <- sapply(kegg.path.cpd.df, function(pl) {
  if (pl == "") return(NA)
  pl.mmu <- intersect(pl, names(path.mouse))
  paste(gsub("path:", "", pl.mmu), collapse = "|")
})
```

## Exporting the results in a SummarizedExperiment object


```{r, creating summarised experiment object for the lung data}

library(SummarizedExperiment)


data.met.lung.varInfo.exp <- data.frame(data.met.lung.varInfo[rownames(dat.met.lung.imp), ],
                                        pval     = mod.lms.pvals,
                                        pval.adj = mod.lms.pvals.adj,
                                        pathway.kegg.id = cross.cpd.id,
                                        pathway.kegg.shortname = as.character(cross.cpd.path),
                                        mod.lm.coefs,
                                        lFC,
                                        lFCl)
# fix strange name in HMDB column
names(data.met.lung.varInfo.exp)[grepl("Group\\.HMDB", names(data.met.lung.varInfo.exp))] <- "Group.HMDB"

se.met.lung <- SummarizedExperiment(assays = list("metab.lung.imp.log" = dat.met.lung.imp.log,
                                                  "metab.lung.raw"     = data.met.lung[rownames(dat.met.lung.imp), ]),
                                    rowData = data.met.lung.varInfo.exp,
                                    colData = meta.data.met.lung)

if (!dir.exists(config$dir.metab)) dir.create(config$dir.metab)
saveRDS(se.met.lung, file = config$file.metab.se)

``` 

## Reproducibility

```{r}
date()
```

```{r}
sessionInfo()
```