---
title: "Lipidomics data analysis"
author: "Sergi Picart"
date: "3/6/2019"
output:
  html_document:
    toc: TRUE
    toc_float: true
    code_folding: hide
    df_print: paged
---

## Summary

This script converts the lung lipidomics dataset we received from Metabolon and saves the relevant information as a SummarizedExperiment object for later use. The main steps include:     
     
* Importing and parsing the data and metadata from the original xls file
* PCA plot of the raw data
* Export results to SummarizedExperiment objects

## Data processing

```{r setup, include=FALSE}
library(pcaMethods)
library(ggplot2)
library(pheatmap)

library(plyr)
library(dplyr)

library(magrittr)

# biohelpeR
library(SummarizedExperiment)
library(MultiAssayExperiment)

knitr::opts_chunk$set(echo = TRUE, message = TRUE)

config <- new.env()
source("config.R", config)

```

### KEGG pathways

```{r}
path.mouse <- KEGGREST::keggList("pathway", organism = "mmu") %>%
  set_names(gsub("path:", "", names(.))) %>%
  gsub(" - Mus.+", "", .)
link.mouse <- KEGGREST::keggLink("pathway", "cpd") %>%
  set_names(gsub("cpd:", "", names(.))) %>%
  gsub("path:map", "mmu", .) %>%
  subset(. %in% names(path.mouse)) %>%
  split(names(.))

list.kegg.pathid <- sapply(link.mouse, paste, collapse = "|")

list.kegg.pathname <- sapply(
  link.mouse, function(cpd) {
    paste(path.mouse[cpd], collapse = "|") 
  }
)
```


### Lung

```{r}
lip.lung <- metabolon2dfs(config$file.lipid.lung, marker.prefix = "LipLung", sample.prefix = "LL_")

# testing metabolon2dfs2() function
# lip.lung <- metabolon2dfs2(file.lipid.lung, 
#                            marker.prefix = "LipLung", 
#                            sample.prefix = "LL_", 
#                            sheet = "ScaledImpData", 
#                            range.rowdata = "A19:G1045", 
#                            range.coldata = "H1:AJ18", 
#                            return.se = F)

# Add the kegg/hmdb mappings
df.aux <- readxl::read_excel(config$file.auxlipid.lung, skip = 7, 
                             sheet = "Pathway Heat Map") %>%
  dplyr::select(`Comp ID`, KEGG, HMDB) %>%
  dplyr::rename(COMP_ID = `Comp ID`) %>% 
  plyr::join(lip.lung$rowdata, by = "COMP_ID", type = "right") %>%
  dplyr::mutate(pathway.kegg.id = list.kegg.pathid[KEGG], 
         pathway.kegg.shortname = list.kegg.pathname[KEGG]) %>%
  set_rownames(.$rownames) 
  

lip.lung$rowdata <- df.aux
rm(df.aux)
```

```{r}
df.hist.lung <- tibble::tibble(original = as.vector(as.matrix(lip.lung$assay)), 
                          log2 = log2(original), 
                          log2plus1 = log2(original + 1)) %>%
  reshape2::melt()

ggplot(df.hist.lung, aes(x = value)) +
  geom_histogram() +
  facet_wrap(~variable, scales = "free_x") +
  theme_bw() +
  theme(aspect.ratio = 1)
```

```{r}
se.lip.lung <- local({
    assay <- as.matrix(lip.lung$assay)
    assay.log <- log2(assay)
    
    coldata <- lip.lung$coldata %>%
      dplyr::mutate(TRT = relevel(factor(TRT), "CTRL"), 
             AGE = relevel(factor(AGE), "3M"))
    
    # fold changes
    grp.trt <- split(rownames(lip.lung$coldata), coldata$TRT)
    grp.age <- split(rownames(lip.lung$coldata), coldata$AGE)
    
    df.fc <- data.frame(
      lFC.AGE = log2(rowMeans(assay[, grp.age$`21M`]) / rowMeans(assay[, grp.age$`3M`])),
      lFC.TRT = log2(rowMeans(assay[, grp.trt$Bleo]) / rowMeans(assay[, grp.trt$CTRL])),
      lFCl.AGE = rowMeans(assay.log[, grp.age$`21M`]) - rowMeans(assay.log[, grp.age$`3M`]),
      lFCl.TRT = rowMeans(assay.log[, grp.trt$Bleo]) - rowMeans(assay.log[, grp.trt$CTRL])
    )
    
    # linear models, coef, anova
    mod.lms <- apply(assay.log, MARGIN = 1, function(metabolite){
                      lm(metabolite ~ coldata$AGE * coldata$TRT)
    })
    
    mod.lms.anova <- lapply(mod.lms, anova)
    mod.lms.pvals <- t(sapply(
      mod.lms.anova, 
      function(mod)
        {matrix(mod[c("coldata$AGE", "coldata$TRT", "coldata$AGE:coldata$TRT"), ]$Pr, ncol = 1)
        }))
    colnames(mod.lms.pvals) <- paste0("pval.", c("AGE", "TRT", "AGE.TRT"))
    mod.lms.pvals.adj <- data.frame(apply(mod.lms.pvals, 2, function(x){p.adjust(x, method = "fdr")}))
    colnames(mod.lms.pvals) <- paste0("pval.adj.", c("AGE", "TRT", "AGE.TRT"))
    
    mod.lm.coefs <- t(sapply(mod.lms, coef))[,-1]
    colnames(mod.lm.coefs) <- paste(
      "coef", 
      sapply(strsplit(colnames(mod.lm.coefs), "$", fixed = TRUE), function(x)
        {if(length(x)>2){"AGE21M.TRTBleo"}else{x[2]}
        }), sep = ".")
    colnames(mod.lm.coefs) <- gsub("Bleo|21M", "", colnames(mod.lm.coefs))

    
    rowdata <- data.frame(
      lip.lung$rowdata, 
      mod.lms.pvals, 
      mod.lm.coefs, 
      df.fc)
  
  
    se.lip.lung <- SummarizedExperiment(assays = list(lip.lung.imp.log = assay.log),
                                      colData = DataFrame(coldata), 
                                      rowData = DataFrame(rowdata))
    
    se.lip.lung
})
```


### Lipid contents

```{r}
if (!dir.exists(config$dir.lipid)) dir.create(config$dir.lipid)

saveRDS(se.lip.lung, 
     file = config$file.lipid.se, 
     compress = "xz")
```


```{r}
se.lipclass.concentration.lung <- metabolon2dfs2(
  config$file.auxlipid.lung, sheet = "Lipid Class Concentrations",
  marker.prefix = "LipClassConLung", 
  sample.prefix = "LCconL_", 
  id.rowname = "CLIENT IDENTIFIER", 
  range.rowdata = "A24:B39", 
  range.coldata = "C6:AE23")

saveRDS(se.lipclass.concentration.lung, 
     file = paste0(config$dir.lipid, "se_lipclass_lung_concentration.rds"), 
     compress = "xz")

se.lipclass.composition.lung <- metabolon2dfs2(
  config$file.auxlipid.lung, sheet = "Lipid Class Compositions",
  marker.prefix = "LipClassComLung", 
  sample.prefix = "LCcomL_", 
  id.rowname = "CLIENT IDENTIFIER", 
  range.rowdata = "A24:B39", 
  range.coldata = "C6:AE23")

saveRDS(se.lipclass.composition.lung, 
     file = paste0(config$dir.lipid, "se_lipclass_lung_composition.rds"), 
     compress = "xz")

se.lipspecies.concentration.lung <- metabolon2dfs2(
  config$file.auxlipid.lung, sheet = "Species Concentrations",
  marker.prefix = "LipSpeciesConLung", 
  sample.prefix = "LSconL_", 
  id.rowname = "CLIENT IDENTIFIER", 
  range.rowdata = "A24:B1040", 
  range.coldata = "C6:AE23")

saveRDS(se.lipspecies.concentration.lung, 
     file = paste0(config$dir.lipid, "se_lipspecies_lung_concentration.rds"), 
     compress = "xz")

se.lipspecies.composition.lung <- metabolon2dfs2(
  config$file.auxlipid.lung, sheet = "Species Compositions",
  marker.prefix = "LipSpeciesComLung", 
  sample.prefix = "LScomL_", 
  id.rowname = "CLIENT IDENTIFIER", 
  range.rowdata = "A24:B1040", 
  range.coldata = "C6:AE23")

saveRDS(se.lipspecies.composition.lung, 
     file = paste0(config$dir.lipid, "se_lipspecies_lung_composition.rds"), 
     compress = "xz")

```


## Descriptive statistics

### Lung


```{r}
summarise_dataset(se.lip.lung, quote.aes = aes(color = AGE, pch = TREATMENT))
```


## Reproducibility

```{r}
date()
```

```{r}
sessionInfo()
```

