---
title: "Bleomycin cross-sectional data formatting"
author: Sergio Picart-Armada
date: "20th May, 2019"
output:
  html_document:
    toc: TRUE
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, error = FALSE, warning = FALSE)
```

## Loading the data

```{r cars}
library(plyr)
library(dplyr)
library(tidyr)
library(magrittr)
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(ggplot2)
library(pcaMethods)

library(data.table)
library(UpSetR)

config <- new.env()
source("config.R", local = config)

# load all the omics data
env.se <- new.env()
local({
  Metabolites <- readRDS(config$file.metab.se)
  Lipids <- readRDS(config$file.lipid.se)
}, env.se)

theme_set(theme_bw())
```

```{r}
# Load the phenotypes
```


## Defining a `MultiAssayExperiment` object

From their vignette, the `MultiAssayExperiment` constructor function can take three arguments:

1. experiments - An ExperimentList or list of data
2. colData - A DataFrame describing the patients (or cell lines, or other biological units)
3. sampleMap - A DataFrame of assay, primary, and colname identifiers

### Experiments

They are already available from the original `rds` files.

### sampleMap: add the same `AnimalID` column to every `colData` 

```{r}
with(env.se, {
  # browser()

  colnames(Metabolites) <- colData(Metabolites)$AnimalID <- paste0(config$id.prefix, 
                                          colData(Metabolites)$ANIMAL_ID)
  
  # rownames(Metabolites) <- rowData(Metabolites)$BIOCHEMICAL

  colnames(Lipids) <- colData(Lipids)$AnimalID <- paste0(config$id.prefix, 
                                          colData(Lipids)$ANIMAL_ID)
  
  # rownames(Lipids) <- rowData(Lipids)$BIOCHEMICAL
})

df.map <- lapply(env.se, function(se) {
  cd <- colData(se)
  data.frame(primary = cd$AnimalID, 
             colname = rownames(cd), 
             stringsAsFactors = FALSE)
}) %>% listToMap
```

Seems like `MultiAssayExperiment` will only take the first assay in every `SummarizedExperiment` object.
Therefore, we must ensure we are picking the preprocessed ones.

```{r}
sapply(env.se, function(se) length(assays(se)))
assays(env.se$Metabolites)
```

In the metabolomics assays, we will get rid of the `raw` data.

```{r}
with(env.se, {
  assays(Metabolites)$metab.lung.raw <- NULL
})
```



### colData: clean and add the phenotypes


### Build the `MultiAssayExperiment` object

Check sample mappings

```{r}
all(colnames(env.se$Lipids) %in% df.map$colname)
all(colnames(env.se$Metabolites) %in% df.map$colname)

all(colData(env.se$Lipids)$AnimalID %in% df.map$primary)
all(colData(env.se$Metabolites)$AnimalID %in% df.map$primary)
all(rownames(df.pheno) %in% df.map$primary)
```

```{r}
setdiff(rownames(df.pheno), df.map$primary)
```

```{r}
# mae <- MultiAssayExperiment(experiments = ExperimentList(as.list(env.se)))
# sampleMap(mae)$primary <- df.map$primary
# # harmonizing input:
# #   removing 56 sampleMap rows with 'primary' not in colData
# #   removing 56 colData rownames not in sampleMap 'primary'
# 
# 
# mae <- MultiAssayExperiment(experiments = ExperimentList(as.list(env.se)), colData = df.pheno)
```

Would the `BIOCHEMICAL` column give proper rownames?
`make.names()` will add periods instead of whitespaces, not ideal...

```{r}
setdiff(make.names(rowData(env.se$Lipids)$BIOCHEMICAL), rowData(env.se$Lipids)$BIOCHEMICAL) %>% head
```


```{r}
# had painful issues with the samples mapping correctly
# specifically, intersectColumns(mae) giving 0 columns
# and the mae having 0 columns in both SEs
# 
# Solution: convert df.pheno to data.frame again and set rownames (it was a DFrame)
mae <- MultiAssayExperiment(experiments = ExperimentList(as.list(env.se)), 
                            colData = df.pheno,
                            sampleMap = df.map)
mae
```

## Basic inspection of the `MultiAssayExperiment` object

As expected, metabolomics has one less sample (28) than transcriptomics and proteomics (29).
Indeed, the 28 samples are common across the datasets.

```{r}
upsetSamples(mae)
```

```{r}
assays(mae)
```

What does the `colData` look like in each omic study?

```{r}
lapply(experiments(mae), function(se) dim(colData(se)))
```

And the `rowData`?

```{r}
lapply(experiments(mae), function(se) dim(rowData(se)))
```

```{r}
if (!dir.exists(config$dir.mae)) dir.create(config$dir.mae)
saveRDS(mae, file = config$out.mae, compress = "xz")
```


## Reproducibility

```{r}
date()
```

```{r}
sessionInfo()
```
