---
title: "Descriptive statistics"
author: Sergio Picart-Armada
date: "20th May, 2019"
output:
  html_document:
    toc: TRUE
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, error = FALSE, warning = FALSE, fig.height = 4, fig.width = 4)
```

## Loading the data

```{r cars}
library(plyr)
library(dplyr)
library(tidyr)
library(magrittr)
library(SummarizedExperiment)
library(MultiAssayExperiment)

library(ggplot2)
library(grid)
library(cowplot)

library(pcaMethods)

library(data.table)
library(UpSetR)

config <- new.env()
source("config.R", local = config)

source("helpers.R")

# load all the omics data
mae <- readRDS(config$out.mae)

if (!dir.exists(config$dir.descriptive)) dir.create(config$dir.descriptive)

theme_set(theme_bw())
```

# Metabolomics

```{r}
se.metab <- experiments(mae)$Metabolites
```

## Descriptive

Number of samples: `r ncol(se.metab)`
Number of features: `r nrow(se.metab)`

Features

```{r}
rowData(se.metab)[, 1:5]
```

How many are not annotated? (those starting with `X - `)

```{r}
grepl("X - ", rowData(se.metab)$BIOCHEMICAL) %>% sum
```

Same ones that don't have a super pathway:

```{r}
sum(is.na(rowData(se.metab)$SUPER_PATHWAY))
```

How many are annotated?

```{r}
grepl("X - ", rowData(se.metab)$BIOCHEMICAL) %>% not %>% sum
```

## PCA


```{r}
gg.metab <- summarise_dataset(
  mae, "Metabolites", 
  quote.aes = aes(shape = AgeTreatment, colour = AgeTreatment), 
  return.interim = TRUE)

# to extract the legend
gg.metab$gg.obj <- gg.metab$gg.obj +
  scale_colour_manual(values = config$col.agetrt, name = "Group") +
  scale_shape_manual(values = config$pch.agetrt, name = "Group") +
  config$ggtheme

ggplot(gg.metab$df.plot, aes(x = PC1, y = PC2, shape = AgeTreatment, colour = AgeTreatment)) +
  geom_hline(yintercept = 0, lty = 2, lwd = .3) +
  geom_vline(xintercept = 0, lty = 2, lwd = .3) +
  geom_point(size = .5) +
  coord_fixed() +
  xlab(gg.metab$names.pc[1]) +
  ylab(gg.metab$names.pc[2]) +
  scale_colour_manual(values = config$col.agetrt.noalpha, name = "Group") +
  scale_shape_manual(values = config$pch.agetrt, name = "Group") +
  scale_x_continuous(limits = get_limits_square(gg.metab$gg.obj)) +
  scale_y_continuous(limits = get_limits_square(gg.metab$gg.obj)) +
  ggtitle(paste0("Metabolites - ", nrow(se.metab), " features")) +
  theme(legend.position = "none") +
  # theme(panel.grid.minor = element_line(size = 0.2), panel.grid.major = element_line(size = .35))
  config$ggtheme


ggsave(filename = paste0(config$dir.descriptive, "/pca_metabolites.svg"), height = config$pca.inches, width = config$pca.inches)
ggsave(filename = paste0(config$dir.descriptive, "/pca_metabolites.png"), height = config$pca.inches, width = config$pca.inches)
ggsave(filename = paste0(config$dir.descriptive, "/pca_metabolites.pdf"), height = config$pca.inches, width = config$pca.inches)
```

Plot legend only

```{r, fig.width=1, fig.height=1.5}
# https://stackoverflow.com/questions/12041042/how-to-plot-just-the-legends-in-ggplot2
gg.legend <- cowplot::get_legend(gg.metab$gg.obj)

grid::grid.newpage()
grid::grid.draw(gg.legend)

ggsave(plot = gg.legend, filename = paste0(config$dir.descriptive, "/pca_legend.svg"), height = 1.5, width = 1)
ggsave(plot = gg.legend, filename = paste0(config$dir.descriptive, "/pca_legend.png"), height = 1.5, width = 1)
ggsave(plot = gg.legend, filename = paste0(config$dir.descriptive, "/pca_legend.pdf"), height = 1.5, width = 1)
```




# Lipidomics

```{r}
se.lipid <- experiments(mae)$Lipids
```

Number of samples: `r ncol(se.lipid)`
Number of features: `r nrow(se.lipid)`

Features:

```{r}
rowData(se.lipid)[, 1:10]
```
How many are not annotated? (those starting with `X - `)

```{r}
grepl("X - ", rowData(se.lipid)$BIOCHEMICAL) %>% sum
```

Same ones that don't have a super pathway:

```{r}
sum(is.na(rowData(se.lipid)$SUPER_PATHWAY))
```

How many are annotated?

```{r}
grepl("X - ", rowData(se.lipid)$BIOCHEMICAL) %>% not %>% sum
```

## PCA


```{r}
gg.lipid <- summarise_dataset(
  mae, "Lipids", 
  quote.aes = aes(shape = Treatment, colour = Age), 
  return.interim = TRUE) 

ggplot(gg.lipid$df.plot, aes(x = PC1, y = PC2, shape = AgeTreatment, colour = AgeTreatment)) +
  geom_hline(yintercept = 0, lty = 2, lwd = .3) +
  geom_vline(xintercept = 0, lty = 2, lwd = .3) +
  geom_point(size = .5) +
  coord_fixed() +
  xlab(gg.lipid$names.pc[1]) +
  ylab(gg.lipid$names.pc[2]) +
  scale_colour_manual(values = config$col.agetrt.noalpha, name = "Group") +
  scale_shape_manual(values = config$pch.agetrt, name = "Group") +
  scale_x_continuous(limits = get_limits_square(gg.lipid$gg.obj)) +
  scale_y_continuous(limits = get_limits_square(gg.lipid$gg.obj)) +
  ggtitle(paste0("Lipids - ", nrow(se.lipid), " features")) +
  theme(legend.position = "none") +
  # theme(panel.grid.minor = element_line(size = 0.2), panel.grid.major = element_line(size = .35))
  config$ggtheme

ggsave(filename = paste0(config$dir.descriptive, "/pca_lipids.svg"), height = config$pca.inches, width = config$pca.inches)
ggsave(filename = paste0(config$dir.descriptive, "/pca_lipids.png"), height = config$pca.inches, width = config$pca.inches)
ggsave(filename = paste0(config$dir.descriptive, "/pca_lipids.pdf"), height = config$pca.inches, width = config$pca.inches)
```


## Reproducibility

```{r}
date()
```

```{r}
sessionInfo()
```
