---
title: "Stratified differential abundance - metabolomics and lipidomics"
author: "Sergio Picart-Armada"
date: "8th March, 2020"
output:
  html_document:
    toc: TRUE
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, 
                      error = FALSE, warning = FALSE, 
                      fig.height = 4)
```

# Loading the data

```{r cars}
library(plyr)
library(dplyr)
library(tidyr)
library(magrittr)

library(SummarizedExperiment)
library(MultiAssayExperiment)

library(ggplot2)
library(ggpmisc)
library(plotly)
# library(GGally)
# library(corrplot)
library(psych)

# library(pls)
# library(pcaMethods)
# library(emmeans)

library(data.table)
# library(UpSetR)

# library(clusterProfiler)
# library(KEGGREST)
# library(pathview)

# metabolomics
# library(FELLA)
# library(igraph)

# library(pheatmap)

config <- new.env()
source("config.R", local = config)
source("helpers.R")

# load all the omics data
mae <- readRDS(config$out.mae)

theme_set(theme_bw())

futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")

if (!dir.exists(config$dir.differential)) dir.create(config$dir.differential)

set.seed(1)
```

## Sanity checks

```{r}
plyr::l_ply(experiments(mae), 
       function(se) stopifnot(levels(colData(se)$TRT)[2] == "Bleo")
)
plyr::l_ply(experiments(mae), 
       function(se) stopifnot(levels(colData(se)$AGE)[2] == "21M")
)

lapply(experiments(mae), 
       function(se) levels(colData(se)$TRT))
lapply(experiments(mae), 
       function(se) levels(colData(se)$AGE))

# check sample mapping concistency
# if this breaks down, check sampleMap(mae)
mae[, colData(mae)$AnimalID]
```

Use chemical names for better readability


```{r}

rownames(experiments(mae)$Lipids) <- rowData(experiments(mae)$Lipids)$BIOCHEMICAL
rownames(experiments(mae)$Metabolites) <- rowData(experiments(mae)$Metabolites)$BIOCHEMICAL

```

# Treatment

## Fit models - treatment

### Classical lm

Markers are guaranteed to lie in logarithmic scale. 
After stratifying by age, they are selected using the model:

$$ y \sim TRT$$

After the models were fit, 

* `log2fc` was computed as the coefficient for `TRTBleo`, 
because the data already lies in log2 scale
* The `p.value`s come from an `anova` (classical lms) or 
`limma::makeContrasts` (using `limma`)
* Multiple test correction: `fdr`


```{r}
# omit the formula argument to use y~x
df.lm <- plyr::ddply(
  as.data.frame(colData(mae)), "Age", function(df.age) {
    mapply(compute_pvals_lfcs, 
      mae[, df.age$AnimalID], 
      var.fc = "TRT", 
      SIMPLIFY = FALSE
    ) %>% plyr::ldply(identity, .id = "se")
  }
) 
```

## Differential entities summary

Significant entities at $FDR < `r config$q.ref`$ and $|\log_2(FC)| > `r config$fc.ref`$ 

### Classical lm

```{r}
mutate(df.lm, signif = spval.adj.TRT < config$q.ref & abs(scoef.TRT) > config$fc.ref) %>%
  group_by(Age, se) %>% 
  summarise(label = paste(sum(signif), length(signif), sep = "/")) %>%
  tidyr::spread(Age, label)
```
### Limma

```{r}
df.limma <- plyr::ddply(
  as.data.frame(colData(mae)), "Age", function(df.age) {
    mapply(limma_pvals_lfcs, 
      mae[, df.age$AnimalID], 
      MoreArgs = list(
        var.fc = "TRT", 
        formula.design = ~TRT, 
        robut = TRUE,
        format.like.lm = TRUE), 
      SIMPLIFY = FALSE
    ) %>% plyr::ldply(identity, .id = "se")
  }
)
```

```{r}
mutate(df.limma, signif = spval.adj.TRT < config$q.ref & abs(scoef.TRT) > config$fc.ref) %>%
  group_by(Age, se) %>% 
  summarise(label = paste(sum(signif), length(signif), sep = "/")) %>%
  tidyr::spread(Age, label)
```

### P-value distributions

Because of the low amount of DE entities in serum, we displayed their p-value distributions.
Regular lm:

```{r, fig.width=5}
ggplot(df.lm, aes(x = spval.TRT)) +
  geom_histogram() +
  facet_grid(Age~se) 
```

Summaries of the q-value distributions:

```{r}
plyr::ddply(df.lm, c("se", "Age"), function(x) summary(na.omit(x$spval.adj.TRT)))
```


Limma:

```{r, fig.width=5}
ggplot(df.limma, aes(x = spval.TRT)) +
  geom_histogram() +
  facet_grid(Age~se)
```

Summaries of the q-value distributions:

```{r}
plyr::ddply(df.limma, c("se", "Age"), function(x) summary(na.omit(x$spval.adj.TRT)))
```


### Final combination

Chose `lm` since `limma` assumes comparable variances across markers.

```{r}
df.fc <- mutate(df.lm, signif = spval.adj.TRT < config$q.ref & abs(scoef.TRT) > config$fc.ref)
write.csv(df.fc, config$df.fc.trt, row.names = FALSE)
```

Number of missing values in p-values

```{r}
sum(is.na(df.fc$spval.adj.TRT))
```

## Statistical checks

Ideas: for each linear model...

* Check normality of residuals: Shapiro-Wilk test on the model residuals
* Check homoscedasticity: F-test on the group variances

P-values must be adjusted since we are performing many tests.
The following function does the dirty work

```{r}
# for each marker...
# fit LM
# get residuals
# compare variances
df.lmdata <- plyr::ddply(
  as.data.frame(colData(mae)), "Age", function(df.age) {
    mapply(compute_statistics_lfcs, 
      mae[, df.age$AnimalID], 
      var.fc = "TRT", 
      SIMPLIFY = FALSE
    ) %>% plyr::ldply(identity, .id = "se")
  }
) 
```

### Shapiro-Wilk test (Normality)

How many markers are significant at FDR < `r config$q.ref`?
Meaning that we reject that their residuals follow a normal distribution.

```{r}
mutate(df.lmdata, signif = fdr.shapiro.TRT < config$q.ref) %>%
  group_by(Age, se) %>% 
  summarise(label = paste(sum(signif, na.rm = TRUE), length(signif), sep = "/")) %>%
  tidyr::spread(Age, label)
```
Overall

```{r}
mutate(df.lmdata, signif = fdr.shapiro.TRT < config$q.ref) %>%
  summarise(Overall = paste(sum(signif, na.rm = TRUE), length(signif), sep = "/"))
```

### Levene's test (homoscedascitity)

See `?car::leveneTest`
Fox, J. (2016) Applied Regression Analysis and Generalized Linear Models, Third Edition. Sage. levene test

```{r}
mutate(df.lmdata, signif = fdr.ltest.TRT < config$q.ref) %>%
  group_by(Age, se) %>% 
  summarise(label = paste(sum(signif, na.rm = TRUE), length(signif), sep = "/")) %>%
  tidyr::spread(Age, label)
```
Overall

```{r}
mutate(df.lmdata, signif = fdr.ltest.TRT < config$q.ref) %>%
  summarise(Overall = paste(sum(signif, na.rm = TRUE), length(signif), sep = "/"))
```

### F-test (Homoscedascitity)

How many markers are significant at FDR < `r config$q.ref`?
Meaning that we reject that the variances of bleo and control are equal.

```{r}
mutate(df.lmdata, signif = fdr.ftest.TRT < config$q.ref) %>%
  group_by(Age, se) %>% 
  summarise(label = paste(sum(signif, na.rm = TRUE), length(signif), sep = "/")) %>%
  tidyr::spread(Age, label)
```

### Breusch-Pagan test (homoscedasticity in linear model)

See `?lmtest::bptest`
T.S. Breusch & A.R. Pagan (1979), A Simple Test for Heteroscedasticity and Random Coefficient Variation. Econometrica 47, 1287–1294
R. Koenker (1981), A Note on Studentizing a Test for Heteroscedasticity. Journal of Econometrics 17, 107–112.

```{r}
mutate(df.lmdata, signif = fdr.bptest.TRT < config$q.ref) %>%
  group_by(Age, se) %>% 
  summarise(label = paste(sum(signif, na.rm = TRUE), length(signif), sep = "/")) %>%
  tidyr::spread(Age, label)
```



# Age

## Fit models 

We will use directly regular lms.

```{r}
# omit the formula argument to use y~x
df.lm.age <- plyr::ddply(
  as.data.frame(colData(mae)), "Treatment", function(df.trt) {
    mapply(compute_pvals_lfcs, 
      mae[, df.trt$AnimalID], 
      var.fc = "AGE", 
      SIMPLIFY = FALSE
    ) %>% plyr::ldply(identity, .id = "se")
  }
) 
```

Significant entities at $FDR < `r config$q.ref`$ and $|\log_2(FC)| > `r config$fc.ref`$ 

```{r}
mutate(df.lm.age, signif = spval.adj.AGE < config$q.ref & abs(scoef.AGE) > config$fc.ref) %>%
  group_by(Treatment, se) %>% 
  summarise(label = paste(sum(signif), length(signif), sep = "/")) %>%
  tidyr::spread(Treatment, label)
```

Number of missing values in p-values

```{r}
sum(is.na(df.lm.age$spval.adj.AGE))
```

### P-value distributions


```{r, fig.width=5}
ggplot(df.lm.age, aes(x = spval.AGE)) +
  geom_histogram() +
  facet_grid(Treatment~se) 
```

Summaries of the q-value distributions:

```{r}
plyr::ddply(df.lm.age, c("se", "Treatment"), function(x) summary(na.omit(x$spval.adj.AGE)))
```

### Export table

Chose `lm`, as for treatment fold changes.

```{r}
df.fc.age <- mutate(df.lm.age, signif = spval.adj.AGE < config$q.ref & abs(scoef.AGE) > config$fc.ref)
write.csv(df.fc.age, config$df.fc.age, row.names = FALSE)
```



# Reproducibility

```{r}
date()
```

```{r}
sessionInfo()
```



